---
title: "LAB Assessment 4"
author: "Lavnish Jhunjhunwala - 24BCE2127"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

# Exercises:
Installing required packages:
``` {r}
install.packages(c("nycflights13", "tidyverse", "moments"))
```

Loading required packages:
``` {r}
library(nycflights13)
library(dplyr)
library(ggplot2)
library(tidyr)
library(moments)
```

---
## 1

### Problem Statement:
    **Arrival Delay Analysis**:
    Calculate the mean, median, and standard deviation for `arr_delay` (arrival delays). How do these compare to the statistics for departure delays calculated earlier?

### Problem Solution:
``` {r}

```



## 2

### Problem Statement:
    **Carrier Variability**:
    Which airline carrier has the most consistent arrival times? Calculate the standard deviation of `arr_delay` for each carrier and identify the one with the lowest value.


### Problem Solution:
``` {r}
```

## 3

### Problem Statement:
    **Speed Distribution**:
    Create a new variable called `speed_mph` by calculating `distance / (air_time / 60)`.
    - Calculate the mean and median speed.
    - Visualize the distribution of flight speeds using a histogram.
    - Is the distribution symmetric or skewed? Calculate the skewness to confirm.

### Problem Solution:
```{r}
```

## 4

### Problem Statement:
    **Interval Data Statistics**:
    Convert the `dep_delay` variable from point data to interval data (grouped data) by creating bins (e.g., using `cut()`).
    - Redo the statistics (mean, standard deviation, and quartiles).
    - Compare the results with the statistics calculated from the original point data.

### Problem Solution:
``` {r}
```


## 5

### Problem Statement:
    **Random Variables and Distributions**:
    Let $X$ be the random variable representing departure delay.
    - Calculate the expected value $E[X]$ and variance $Var(X)$ from the data.
    - Visualize the probability density function (PDF) of $X$.
    - Based on the shape, skewness, and kurtosis, discuss which theoretical distribution might fit the data.

### Problem Solution:
```{r}
```


## 6

### Problem Statement:
    **On-time Performance Test**:
    Conduct a hypothesis test to check if there is an association between the origin airport (`origin`) and whether a flight arrives on time (within 5 minutes of scheduled arrival).
    - Create a contingency table.
    - Run a Chi-squared test.
    - Interpret the results.

### Problem Solution:
```{r}
```