---
title: "Module 1: Introduction to Statistics"
author: "Statistics Course"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction to Statistics

## Overview
In this module, we explore fundamental statistical concepts including:
- Statistics and Data Analysis
- Measures of Central Tendency
- Measures of Dispersion
- Moments, Skewness, and Kurtosis

We'll use the `nycflights13` dataset to demonstrate these concepts with real-world data.

---

## Data Preparation

First, let's load the necessary libraries and explore the NYC flights dataset:

```{r load_data}
# Install packages if needed (uncomment to run)
# install.packages(c("nycflights13", "tidyverse", "moments"))

library(nycflights13)
library(dplyr)
library(ggplot2)
library(tidyr)
library(moments)

# Load the flights dataset
data(flights)

# Display basic information
head(flights)
cat("\nDataset dimensions:", nrow(flights), "rows x", ncol(flights), "columns\n")
cat("\nColumn names:\n")
print(names(flights))
```

---

## 1. Statistics and Data Analysis

Statistics involves collecting, analyzing, and interpreting data to understand patterns and relationships.

### What is Statistical Data Analysis?

Statistical data analysis helps us:
- Describe the characteristics of data
- Make inferences about populations
- Test hypotheses
- Predict future trends

### Example: Airline Delay Analysis

Let's analyze departure delays across different airlines:

```{r data_analysis}
# Create a summary of airlines in the dataset
airline_summary <- flights %>%
  group_by(carrier) %>%
  summarise(
    n_flights = n(),
    avg_delay = mean(dep_delay, na.rm = TRUE),
    max_delay = max(dep_delay, na.rm = TRUE),
    min_delay = min(dep_delay, na.rm = TRUE)
  ) %>%
  arrange(desc(avg_delay))

print(airline_summary)

# Visualize average delays by airline
ggplot(airline_summary, aes(x = reorder(carrier, avg_delay), y = avg_delay)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Average Departure Delay by Airline",
       x = "Airline",
       y = "Average Delay (minutes)") +
  theme_minimal()
```

---

## 2. Measures of Central Tendency

Central tendency measures describe the "center" or typical value of a dataset.

### Types of Central Tendency

**2.1 Mean (Average)**

The arithmetic mean is the sum of all values divided by the count of values.

$$\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i$$

```{r mean}
# Calculate mean departure delay
mean_delay <- mean(flights$dep_delay, na.rm = TRUE)
cat("Mean departure delay:", round(mean_delay, 2), "minutes\n")

# By month
mean_by_month <- flights %>%
  group_by(month) %>%
  summarise(mean_delay = mean(dep_delay, na.rm = TRUE),
            .groups = 'drop')

print(mean_by_month)

# Visualize mean delays by month
ggplot(mean_by_month, aes(x = month, y = mean_delay)) +
  geom_line(color = "darkblue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(title = "Mean Departure Delay by Month",
       x = "Month",
       y = "Mean Delay (minutes)") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:12)
```

**2.2 Median (Middle Value)**

The median is the middle value when data is sorted. It divides the dataset into two equal halves.

```{r median}
# Calculate median departure delay
median_delay <- median(flights$dep_delay, na.rm = TRUE)
cat("Median departure delay:", median_delay, "minutes\n")

# Compare mean vs median for each month
comparison <- flights %>%
  group_by(month) %>%
  summarise(
    mean_delay = mean(dep_delay, na.rm = TRUE),
    median_delay = median(dep_delay, na.rm = TRUE),
    .groups = 'drop'
  )

print(comparison)

# Visualize comparison
comparison_long <- comparison %>%
  pivot_longer(cols = c(mean_delay, median_delay),
               names_to = "measure",
               values_to = "delay")

ggplot(comparison_long, aes(x = month, y = delay, color = measure)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Mean vs Median Departure Delay by Month",
       x = "Month",
       y = "Delay (minutes)",
       color = "Measure") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:12)
```

**2.3 Mode (Most Frequent Value)**

The mode is the value that appears most frequently in the dataset.

```{r mode}
# Calculate mode
calculate_mode <- function(x) {
  freq <- table(x)
  as.numeric(names(freq)[which.max(freq)])
}

# Mode is often used for categorical data
mode_origin <- calculate_mode(flights$origin)
cat("Mode of origin:", mode_origin, "\n")

# Frequency of origins
origin_freq <- flights %>%
  group_by(origin) %>%
  summarise(count = n(), .groups = 'drop') %>%
  arrange(desc(count))

print(origin_freq)

# Visualize
ggplot(origin_freq, aes(x = reorder(origin, -count), y = count)) +
  geom_col(fill = "darkgreen") +
  labs(title = "Frequency of Flight Origins",
       x = "Airport",
       y = "Number of Flights") +
  theme_minimal()
```

**Comparison of Central Tendency Measures:**

```{r central_tendency_comparison}
# Overall comparison
cat("=== Central Tendency Measures for Departure Delay ===\n")
cat("Mean:  ", round(mean(flights$dep_delay, na.rm = TRUE), 2), "minutes\n")
cat("Median:", median(flights$dep_delay, na.rm = TRUE), "minutes\n")
cat("Mode:  ", calculate_mode(round(flights$dep_delay)), "minutes\n")

# Summary statistics
cat("\n=== Summary Statistics ===\n")
print(summary(flights$dep_delay))
```

---

## 3. Measures of Dispersion

Dispersion measures describe the spread or variability of data around the center.

### 3.1 Range

Range is the difference between the maximum and minimum values.

$$\text{Range} = \max(x) - \min(x)$$

```{r range}
range_delay <- max(flights$dep_delay, na.rm = TRUE) - min(flights$dep_delay, na.rm = TRUE)
cat("Range of departure delay:", range_delay, "minutes\n")
cat("Min delay:", min(flights$dep_delay, na.rm = TRUE), "minutes\n")
cat("Max delay:", max(flights$dep_delay, na.rm = TRUE), "minutes\n")
```

### 3.2 Variance

Variance measures the average squared deviation from the mean.

$$s^2 = \frac{1}{n}\sum_{i=1}^{n} (x_i - \bar{x})^2$$

```{r variance}
var_delay <- var(flights$dep_delay, na.rm = TRUE)
cat("Variance of departure delay:", round(var_delay, 2), "\n")

# By airline
var_by_airline <- flights %>%
  group_by(carrier) %>%
  summarise(
    variance = var(dep_delay, na.rm = TRUE),
    n = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(variance))

print(var_by_airline)
```

### 3.3 Standard Deviation

Standard deviation is the square root of variance, expressing dispersion in original units.

$$s = \sqrt{\frac{1}{n}\sum_{i=1}^{n} (x_i - \bar{x})^2}$$

```{r standard_deviation}
sd_delay <- sd(flights$dep_delay, na.rm = TRUE)
cat("Standard deviation of departure delay:", round(sd_delay, 2), "minutes\n")

# By month
sd_by_month <- flights %>%
  group_by(month) %>%
  summarise(
    sd_delay = sd(dep_delay, na.rm = TRUE),
    mean_delay = mean(dep_delay, na.rm = TRUE),
    .groups = 'drop'
  )

print(sd_by_month)

# Visualize with confidence bands
ggplot(sd_by_month, aes(x = month, y = mean_delay)) +
  geom_line(color = "darkblue", size = 1) +
  geom_point(color = "red", size = 2) +
  geom_ribbon(aes(ymin = mean_delay - sd_delay, ymax = mean_delay + sd_delay),
              alpha = 0.2, fill = "blue") +
  labs(title = "Mean Departure Delay with Standard Deviation Band",
       x = "Month",
       y = "Delay (minutes)") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:12)
```

### 3.4 Quartiles and Interquartile Range (IQR)

Quartiles divide the data into four equal parts.

```{r quartiles}
# Calculate quartiles
q1 <- quantile(flights$dep_delay, 0.25, na.rm = TRUE)
q2 <- quantile(flights$dep_delay, 0.50, na.rm = TRUE)  # Median
q3 <- quantile(flights$dep_delay, 0.75, na.rm = TRUE)
iqr <- q3 - q1

cat("Q1 (25th percentile):", q1, "\n")
cat("Q2 (50th percentile/Median):", q2, "\n")
cat("Q3 (75th percentile):", q3, "\n")
cat("Interquartile Range (IQR):", iqr, "\n")

# Box plot by airline
ggplot(flights, aes(x = carrier, y = dep_delay)) +
  geom_boxplot(fill = "lightblue", outlier.alpha = 0.3) +
  labs(title = "Distribution of Departure Delays by Airline",
       x = "Airline",
       y = "Departure Delay (minutes)") +
  theme_minimal() +
  ylim(-20, 150)  # Limit y-axis for better visualization
```

### 3.5 Coefficient of Variation

CV measures relative variability as a percentage of the mean.

$$CV = \frac{s}{\bar{x}} \times 100\%$$

```{r cv}
mean_delay <- mean(flights$dep_delay, na.rm = TRUE)
sd_delay <- sd(flights$dep_delay, na.rm = TRUE)
cv <- (sd_delay / mean_delay) * 100

cat("Coefficient of Variation:", round(cv, 2), "%\n")

# By airline
cv_by_airline <- flights %>%
  group_by(carrier) %>%
  summarise(
    mean = mean(dep_delay, na.rm = TRUE),
    sd = sd(dep_delay, na.rm = TRUE),
    cv = (sd / mean) * 100,
    .groups = 'drop'
  ) %>%
  arrange(desc(cv))

print(cv_by_airline)
```

---

## 4. Moments, Skewness, and Kurtosis

### 4.1 Moments

Moments describe the shape and characteristics of a distribution.

**Raw Moments:** $m_r = \frac{1}{n}\sum_{i=1}^{n} (x_i)^r$

**Central Moments:** $\mu_r = \frac{1}{n}\sum_{i=1}^{n} (x_i - \bar{x})^r$

```{r moments}
# Calculate moments
m1 <- mean(flights$dep_delay, na.rm = TRUE)
m2 <- mean((flights$dep_delay - m1)^2, na.rm = TRUE)
m3 <- mean((flights$dep_delay - m1)^3, na.rm = TRUE)
m4 <- mean((flights$dep_delay - m1)^4, na.rm = TRUE)

cat("First moment (Mean):", round(m1, 2), "\n")
cat("Second moment (Variance):", round(m2, 2), "\n")
cat("Third moment:", round(m3, 2), "\n")
cat("Fourth moment:", round(m4, 2), "\n")
```

### 4.2 Skewness

Skewness measures the asymmetry of the distribution.

$$\text{Skewness} = \frac{\mu_3}{s^3}$$

- **Skewness = 0**: Symmetric distribution
- **Skewness > 0**: Right-skewed (positively skewed)
- **Skewness < 0**: Left-skewed (negatively skewed)

```{r skewness}
# Calculate skewness
skewness_value <- skewness(flights$dep_delay, na.rm = TRUE)
cat("Skewness of departure delay:", round(skewness_value, 3), "\n")

if (skewness_value > 0.5) {
  cat("Interpretation: Right-skewed (positive skew) distribution\n")
} else if (skewness_value < -0.5) {
  cat("Interpretation: Left-skewed (negative skew) distribution\n")
} else {
  cat("Interpretation: Approximately symmetric distribution\n")
}

# Skewness by month
skewness_by_month <- flights %>%
  group_by(month) %>%
  summarise(
    skewness = skewness(dep_delay, na.rm = TRUE),
    .groups = 'drop'
  )

print(skewness_by_month)

# Visualize
ggplot(skewness_by_month, aes(x = month, y = skewness)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Skewness of Departure Delays by Month",
       x = "Month",
       y = "Skewness") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:12)

# Distribution histogram with density curve
ggplot(flights, aes(x = dep_delay)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "lightblue", color = "black") +
  geom_density(color = "red", size = 1) +
  labs(title = "Distribution of Departure Delays (Showing Skewness)",
       x = "Departure Delay (minutes)",
       y = "Density") +
  theme_minimal() +
  xlim(-20, 200)
```

### 4.3 Kurtosis

Kurtosis measures the "tailedness" or peakedness of the distribution.

$$\text{Kurtosis} = \frac{\mu_4}{s^4} - 3$$

- **Excess Kurtosis = 0**: Mesokurtic (Normal distribution)
- **Excess Kurtosis > 0**: Leptokurtic (Heavy tails, sharp peak)
- **Excess Kurtosis < 0**: Platykurtic (Light tails, flat peak)

```{r kurtosis}
# Calculate kurtosis (excess kurtosis)
kurtosis_value <- kurtosis(flights$dep_delay, na.rm = TRUE)
cat("Excess Kurtosis of departure delay:", round(kurtosis_value, 3), "\n")

if (kurtosis_value > 0) {
  cat("Interpretation: Leptokurtic - Heavy tails with sharp peak\n")
} else if (kurtosis_value < 0) {
  cat("Interpretation: Platykurtic - Light tails with flat peak\n")
} else {
  cat("Interpretation: Mesokurtic - Similar to normal distribution\n")
}

# Kurtosis by month
kurtosis_by_month <- flights %>%
  group_by(month) %>%
  summarise(
    kurtosis = kurtosis(dep_delay, na.rm = TRUE),
    .groups = 'drop'
  )

print(kurtosis_by_month)

# Visualize
ggplot(kurtosis_by_month, aes(x = month, y = kurtosis)) +
  geom_col(fill = "darkgreen") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Kurtosis of Departure Delays by Month",
       x = "Month",
       y = "Excess Kurtosis") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:12)
```

---

## 5. Hypothesis Test: On-time departures predict on-time arrivals

  We test the hypothesis: "Flights that left within 5 minutes of their scheduled departure arrived within 5 minutes of their scheduled arrival." Below we filter the data, create logical indicators for leaving/arriving within ±5 minutes, show contingency counts and proportions, run an appropriate test of association (chi-squared or Fisher's exact), compute conditional probabilities, plot the results, and provide a brief interpretation.

  ```{r hypothesis-test, echo=TRUE, message=FALSE, warning=FALSE}
  library(dplyr)
  library(ggplot2)

  # Filter and create indicators
  fl_ht <- flights %>%
    filter(!is.na(dep_delay), !is.na(arr_delay)) %>%
    mutate(
      left_on_time = abs(dep_delay) <= 5,
      arrived_on_time = abs(arr_delay) <= 5
    )

  # Contingency table and row proportions
  tab <- table(Left = fl_ht$left_on_time, Arrived = fl_ht$arrived_on_time)
  tab

  prop_tab <- prop.table(tab, margin = 1)
  round(prop_tab, 3)

  # Statistical test: chi-squared if expected counts are large enough, otherwise Fisher's exact
  # Use suppressWarnings when computing expected to avoid extra output if assumptions borderline
  expected <- suppressWarnings(chisq.test(tab))$expected
  use_fisher <- any(expected < 5)

  if (use_fisher) {
    test_result <- fisher.test(tab)
  } else {
    test_result <- chisq.test(tab)
  }
  test_result

  # Conditional probabilities
  P_arrive_if_left <- prop_tab["TRUE", "TRUE"]
  P_arrive_if_not_left <- prop_tab["FALSE", "TRUE"]
  cat(sprintf("P(arrived_on_time | left_on_time) = %.3f\n", P_arrive_if_left))
  cat(sprintf("P(arrived_on_time | NOT left_on_time) = %.3f\n", P_arrive_if_not_left))

  # Visualization: proportion arrived on time by whether flight left on time
  viz_df <- fl_ht %>%
    group_by(left_on_time) %>%
    summarize(n = n(), prop_arrived_on_time = mean(arrived_on_time), .groups = "drop")

  ggplot(viz_df, aes(x = factor(left_on_time), y = prop_arrived_on_time, fill = factor(left_on_time))) +
    geom_col() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(x = "Left within 5 min of scheduled departure?", y = "Proportion arrived within 5 min",
         title = "Arrival punctuality by departure punctuality") +
    theme_minimal() +
    scale_x_discrete(labels = c("FALSE" = "No", "TRUE" = "Yes")) +
    guides(fill = "none")

  ```

  **Interpretation (how to read results):**

  - The contingency table shows counts of flights that arrived on time or not, split by whether they left on time.
  - The chi-squared (or Fisher) test reports whether there is evidence of association between departure punctuality and arrival punctuality. A small p-value (e.g., < 0.05) indicates an association.
  - The conditional probabilities P(arrived_on_time | left_on_time) and P(arrived_on_time | NOT left_on_time) give practical measures of how predictive departure punctuality is for arrival punctuality.

  Note: This is observational data; association does not imply causation.

---

## 6. Summary: Complete Statistical Profile

Let's create a comprehensive summary of all statistical measures:

```{r summary_stats}
# Comprehensive statistical summary
summary_stats <- data.frame(
  Measure = c("Mean", "Median", "Std Dev", "Variance", "Range", "Q1", "Q3", "IQR", 
              "Skewness", "Kurtosis", "CV (%)"),
  Value = c(
    round(mean(flights$dep_delay, na.rm = TRUE), 2),
    median(flights$dep_delay, na.rm = TRUE),
    round(sd(flights$dep_delay, na.rm = TRUE), 2),
    round(var(flights$dep_delay, na.rm = TRUE), 2),
    max(flights$dep_delay, na.rm = TRUE) - min(flights$dep_delay, na.rm = TRUE),
    quantile(flights$dep_delay, 0.25, na.rm = TRUE),
    quantile(flights$dep_delay, 0.75, na.rm = TRUE),
    quantile(flights$dep_delay, 0.75, na.rm = TRUE) - quantile(flights$dep_delay, 0.25, na.rm = TRUE),
    round(skewness(flights$dep_delay, na.rm = TRUE), 3),
    round(kurtosis(flights$dep_delay, na.rm = TRUE), 3),
    round((sd(flights$dep_delay, na.rm = TRUE) / mean(flights$dep_delay, na.rm = TRUE)) * 100, 2)
  )
)

cat("=== COMPLETE STATISTICAL SUMMARY ===\n")
print(summary_stats)

# Export summary
write.csv(summary_stats, "output/statistical_summary.csv", row.names = FALSE)
cat("\nSummary saved to output/statistical_summary.csv\n")
```

---

## Conclusion

This module has covered the fundamental concepts of statistics:

1. **Statistics and Data Analysis** - Understanding data patterns and relationships
2. **Measures of Central Tendency** - Mean, Median, and Mode describe typical values
3. **Measures of Dispersion** - Range, Variance, SD, and IQR describe spread
4. **Moments, Skewness, and Kurtosis** - Describe distribution shape and characteristics

These tools form the foundation for more advanced statistical methods and hypothesis testing.

---

## Exercises

1.  **Arrival Delay Analysis**:
    Calculate the mean, median, and standard deviation for `arr_delay` (arrival delays). How do these compare to the statistics for departure delays calculated earlier?

2.  **Carrier Variability**:
    Which airline carrier has the most consistent arrival times? Calculate the standard deviation of `arr_delay` for each carrier and identify the one with the lowest value.

3.  **Speed Distribution**:
    Create a new variable called `speed_mph` by calculating `distance / (air_time / 60)`.
    - Calculate the mean and median speed.
    - Visualize the distribution of flight speeds using a histogram.
    - Is the distribution symmetric or skewed? Calculate the skewness to confirm.

4.  **Interval Data Statistics**:
    Convert the `dep_delay` variable from point data to interval data (grouped data) by creating bins (e.g., using `cut()`).
    - Redo the statistics (mean, standard deviation, and quartiles).
    - Compare the results with the statistics calculated from the original point data.

5.  **Random Variables and Distributions**:
    Let $X$ be the random variable representing departure delay.
    - Calculate the expected value $E[X]$ and variance $Var(X)$ from the data.
    - Visualize the probability density function (PDF) of $X$.
    - Based on the shape, skewness, and kurtosis, discuss which theoretical distribution might fit the data.

6.  **On-time Performance Test**:
    Conduct a hypothesis test to check if there is an association between the origin airport (`origin`) and whether a flight arrives on time (within 5 minutes of scheduled arrival).
    - Create a contingency table.
    - Run a Chi-squared test.
    - Interpret the results.

---

## References

- Wickham, H. (2016). ggplot2: Elegant Graphics for Data Analysis.
- James, G., Witten, D., Hastie, T., & Tibshirani, R. (2013). An Introduction to Statistical Learning.
- nycflights13 R Package: https://github.com/hadley/nycflights13
